---
title: "11-20 Milestone"
output: html_document
date: "2025-11-17"
---

# Project Description

### Why did we choose this dataset (motivation)?

### Describe the dataset

## Data Cleaning

```{r}
summary(cars)
```

## Coming up with our research questions

```{r}
plot(pressure)
```

## [Individual] Explain why you chose your research question and what you plan to do with it

### Gian

my main research questions have to do with spatial data within the city of Boston. Since the dataset includes zipcode and address for each restaurant, I thought it would be interesting to create a map that included each subsectors of the city (by zipcode) and investigate the differences and similarities between them.

I may also be able to do geospatial computations that include distance between restaurants and other restaurants or reference points. This will all be done using the **sf** package.

My main visualizations will all be based on the map of Boston with the zipcode subsectors. I would like to do a density map with the amount of health code violations, as well as a map that categorizes each zipcode by the most common type of violation in for each zipcode (Dishwashing Facilities, Signage, Cross Contamination, Building Maintenance, etc) by color.

### Niki

### Seun

### Arohi
