---
title: "11-20 Milestone"
output: html_document
date: "2025-11-17"
---

# Project Description

### Why did we choose this dataset (motivation)?

### Describe the dataset

## Data Cleaning

```{r}
summary(cars)
```

## Coming up with our research questions

```{r}
plot(pressure)
```

## EDA questions to work on:
1. What is the time range of the data? Are there any gaps, peaks, or dips?
Output: Histogram of dates
Why: To show periods of data collection, patterns in how it is collected, and if there are any gaps in the data collection process. 

2. What is the distribution of inspection outcomes and risk levels across all cities?
Output: Bar Chart of Risk; Bar Chart of Outcomes
Why: To show the overall ratio of "High" vs. "Low" risk violations and "Pass" vs. "Fail" in the entire dataset

3. For cities with scores (NYC, SF), what does the score distribution look like?
Output: Histogram of scores
Why: To see the range of grades given, and where most restaurants fall or cluster around

4. Which cities have the highest proportion of "Fail" outcomes or "High Risk" violations?
Output: Bar Chart

## [Individual] Explain why you chose your research question and what you plan to do with it

### Gian

my main research questions have to do with spatial data within the city of Boston. Since the dataset includes zipcode and address for each restaurant, I thought it would be interesting to create a map that included each subsectors of the city (by zipcode) and investigate the differences and similarities between them.

I may also be able to do geospatial computations that include distance between restaurants and other restaurants or reference points. This will all be done using the **sf** package.

My main visualizations will all be based on the map of Boston with the zipcode subsectors. I would like to do a density map with the amount of health code violations, as well as a map that categorizes each zipcode by the most common type of violation in for each zipcode (Dishwashing Facilities, Signage, Cross Contamination, Building Maintenance, etc) by color.

### Niki

### Seun

### Arohi
